---
title: "p8105_hw2_yl6377"
output:
  html_document: default
  pdf_document: default
date: "2026-10-05"
---


```{r}
library(tidyverse)
```
Problem 1
# read and clean data
```{r}
nyc_transit =
  read_csv(
    "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("","NA",".")
  )
nyc_transit = nyc_transit |>
  select(
    line = `Line`,
    division = `Division`,
    name = `Station Name`,
    station_latitude = `Station Latitude`,
    station_longitude = `Station Longitude`,
    route1 = `Route1`,
    route2 = `Route2`,
    route3 = `Route3`,
    route4 = `Route4`,
    route5 = `Route5`,
    route6 = `Route6`,
    route7 = `Route7`,
    route8 = `Route8`,
    route9 = `Route9`,
    route10 = `Route10`,
    route11 = `Route11`,
    entry = `Entry`,
    vending = `Vending`,
    entrance_type = `Entrance Type`,
    ada = `ADA`
  ) |>
  mutate(
    entry=case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```
The NYC Transit dataset contains information about subway stations and their entrances and exits, including the subway line, station name, station latitude and longitude, routes served, whether an entrance allows entry, vending availability, entrance type, and ADA compliance. I cleaned the dataset by retaining the variables relevant to the analysis and converting the entry variable from character values (YES/NO) to a logical variable (TRUE/FALSE). The resulting dataset has `r nrow(nyc_transit)` rows and `r ncol(nyc_transit)` columns. The dimensions of the dataset are `r nrow(nyc_transit)` rows × `r ncol(nyc_transit)` columns. These data are not completely tidy because the routes served by each station are stored across multiple variables (Route1 through Route8) rather than with one route per observation.

1.1 How many distinct stations?
```{r}
nyc_transit |>
  distinct(division, name) |>
  nrow()
```

1.2 How many stations are ADA compliant?
```{r}
nyc_transit |>
  distinct(division, name, ada) |>
  filter(ada == TRUE) |>
  nrow()
```

1.3 What proportion of entrances/exits without vending allow entrance?
```{r}
nyc_transit |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry, na.rm = TRUE)
  )
```

```{r}
routes= 
nyc_transit |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )|>
  filter(!is.na(route_name))
```

1.4 How many distinct stations serve the A train?
```{r}
route_A =
  routes |>
  filter(route_name == "A") |>
  distinct(line, name)|>
  nrow()

route_A
```

1.5 Of the stations that serve the A train, how many are ADA compliant?
```{r}
route_A_ada =
  routes |>
  filter(route_name == "A") |>
  distinct(line, name,ada) |>
  filter(ada == TRUE) |>
  nrow()

route_A_ada
```
