---
title: "p8105_hw2_yl6377"
output:
  html_document: default
  pdf_document: default
date: "2026-10-05"
---


```{r}
library(tidyverse)
library(readxl)
```

Problem 1
# read and clean NYC Transit data
```{r}
nyc_transit =
  read_csv(
    "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    na = c("","NA",".")
  )|>
  select(
    line = `Line`,
    division = `Division`,
    name = `Station Name`,
    station_latitude = `Station Latitude`,
    station_longitude = `Station Longitude`,
    route1 = `Route1`,
    route2 = `Route2`,
    route3 = `Route3`,
    route4 = `Route4`,
    route5 = `Route5`,
    route6 = `Route6`,
    route7 = `Route7`,
    route8 = `Route8`,
    route9 = `Route9`,
    route10 = `Route10`,
    route11 = `Route11`,
    entry = `Entry`,
    vending = `Vending`,
    entrance_type = `Entrance Type`,
    ada = `ADA`
  ) |>
  mutate(
    entry=case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )
```
The NYC Transit dataset contains information about subway stations and their entrances and exits, including the subway line, station name, station latitude and longitude, routes served, whether an entrance allows entry, vending availability, entrance type, and ADA compliance. I cleaned the dataset by retaining the variables relevant to the analysis and converting the entry variable from character values (YES/NO) to a logical variable (TRUE/FALSE). The resulting dataset has `r nrow(nyc_transit)` rows and `r ncol(nyc_transit)` columns. The dimensions of the dataset are `r nrow(nyc_transit)` rows × `r ncol(nyc_transit)` columns. These data are not completely tidy because the routes served by each station are stored across multiple variables (Route1 through Route8) rather than with one route per observation.

1.1 How many distinct stations?
```{r}
nyc_transit |>
  distinct(division, name) |>
  nrow()
```

1.2 How many stations are ADA compliant?
```{r}
nyc_transit |>
  distinct(division, name, ada) |>
  filter(ada == TRUE) |>
  nrow()
```

1.3 What proportion of entrances/exits without vending allow entrance?
```{r}
nyc_transit |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry, na.rm = TRUE)
  )
```

```{r}
routes= 
nyc_transit |>
  mutate(
    across(starts_with("route"), as.character)
  ) |>
  pivot_longer(
    cols = starts_with("route"),
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  )|>
  filter(!is.na(route_name))
```

1.4 How many distinct stations serve the A train?
```{r}
route_A =
  routes |>
  filter(route_name == "A") |>
  distinct(line, name)|>
  nrow()

route_A
```

1.5 Of the stations that serve the A train, how many are ADA compliant?
```{r}
route_A_ada =
  routes |>
  filter(route_name == "A") |>
  distinct(line, name,ada) |>
  filter(ada == TRUE) |>
  nrow()

route_A_ada
```

Problem 2
# data analysis of trash wheel

Mr. Trash Wheel Sheet
```{r}
mr_trash_wheel =
  read_excel(
    "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    skip = 1
  )|>
  janitor::clean_names()|>
  select(1:14) |>
  drop_na(dumpster) |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )
```

Professor Trash Wheel Sheet
```{r}
professor_trash_wheel =
  read_excel(
    "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    skip=1
  )|>
  janitor::clean_names()|>
  drop_na(dumpster) |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash Wheel"
  )
```

Gwynnda the Good Wheel of the W
```{r}
gwynnda =
  read_excel(
    "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    skip = 1
  )|>
  janitor::clean_names()|>
  drop_na(dumpster) |>
  mutate(
    year = as.numeric(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Gwynnda"
  )
```

Combine the three datasets
```{r}
all_trash_wheel =
  bind_rows(
    mr_trash_wheel,
    professor_trash_wheel,
    gwynnda
  )
```

2.1 What was the total weight of trash collected by Professor Trash Wheel? 
```{r}
professor_total_weight =
all_trash_wheel |>
  filter(trash_wheel == "Professor Trash Wheel") |>
  summarise(
    total_weight = sum(weight_tons, na.rm = TRUE)
  )
```

2.2 What was the total number of cigarette butts collected by Gwynnda in June of 2022?
```{r}
gwynnda_june_2022 =
all_trash_wheel |>
  filter(
    trash_wheel == "Gwynnda",
    year == 2022,
    str_trim(month) == "June"
  ) |>
  summarise(
    total_cigarette_butts = sum(cigarette_butts, na.rm = TRUE)
  )
```

The resulting dataset contains `r nrow(all_trash_wheel)` observations from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Key variables include `r names(all_trash_wheel)`. For the available data, Professor Trash Wheel collected a total of `r professor_total_weight` tons of trash. Gwynnda collected a total of `r gwynnda_june_2022` cigarette butts in June 2022.

Problem 3
# data analysis of AD biomarkers observation

Import, clean, and tidy baseline demographics
```{r}
baseline_raw =
  read_csv(
    "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/data_mci/MCI_baseline.csv",
    skip = 1,
    na = c(".", "Na", "")
  ) |>
  janitor::clean_names() 
```

```{r}
baseline =
  baseline_raw |>
  mutate(
    sex = factor(
      sex,
      levels = c(0, 1),
      labels = c("Female", "Male")
    ),
    apoe4 = factor(
      apoe4,
      levels = c(0, 1),
      labels = c("Non-carrier", "Carrier")
    )
  ) |>
  filter(
    is.na(age_at_onset) | age_at_onset > current_age
  )
```

3.1 How many participants were recruited?
```{r}
  nrow(baseline_raw)
```

3.2 How many are eligible after inclusion criterion?
```{r}
  nrow(baseline)
```

3.3 Of eligible participants, how many developed MCI?
```{r}
  sum(!is.na(baseline$age_at_onset))
```

3.4 The average baseline age is 
```{r}
  mean(baseline$current_age, na.rm = TRUE)
```

3.5 Proportion of women who are APOE4 carriers
```{r}
baseline |>
  filter(sex == "Female") |>
  summarise(
    proportion = mean(apoe4 == "Carrier", na.rm = TRUE)
  )
```

3.6 Import and tidy longitudinal amyloid data
```{r}
amyloid =
  read_csv(
    "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/data_mci//mci_amyloid.csv",
    skip = 1,
    na = c("Na", "")
  ) |>
  janitor::clean_names() |>
  rename(
    id = study_id
  ) |>
  pivot_longer(
    cols = -id,
    names_to = "visit",
    values_to = "amyloid_42_40"
  ) |>
  mutate(
    amyloid_42_40 = as.numeric(amyloid_42_40)
  )
```
3.7 Check participants only in amyloid
```{r}
amyloid_only =
  amyloid |>
  distinct(id) |>
  anti_join(
    baseline,
    by = "id"
  )

nrow(amyloid_only)
```

3.8 Find participants in both datasets
```{r}
intersect(
  baseline$id,
  unique(amyloid$id)
) |>
  length()
```

3.9 Combine the two datasets
```{r}
all_mci =
  inner_join(
    baseline,
    amyloid,
    by = "id"
  )
```

3.10 Export the final dataset
```{r}
write_csv(
  all_mci,
  "/Users/anna/Desktop/CU/Data Science I/p8105_hw2_yl6377/data_mci/mci_result.csv"
)
```

Description:
A total of `r nrow(baseline_raw)` participants were recruited. After applying the
inclusion criterion that participants were free of MCI at baseline,
`r nrow(baseline)` participants remained. Of these eligible participants,
`r sum(!is.na(baseline$age_at_onset))` developed MCI during follow-up.
The average baseline age was `r round(mean(baseline$current_age, na.rm = TRUE), 2)` years.
Among women, `r mean(baseline$apoe4[baseline$sex == "Female"] == "Carrier", na.rm = TRUE) * 100`% were APOE4 carriers.
The two datasets were combined using an inner join, retaining `r n_distinct(all_mci$id)` participants who were present in both datasets. The final dataset contains `r nrow(all_mci)` observations and `r ncol(all_mci)` variables.